Hydras \& Co.: Formalized mathematics in Coq\\ for inspiration and entertainment

### 2022-04-25 doc/JFLA2022/Slides/Slides.tex
work on JFLA slides

\documentclass[10pt]{beamer}
\usetheme{Montpellier}
\usepackage{xspace, alltt}
\usepackage[utf8x]{inputenc}

\usepackage{fontspec}
%\setmainfont{DejaVu Serif}

\setmonofont{DejaVu Sans Mono}[Scale=MatchLowercase]

\newcommand{\TODO}[2][]{[\textcolor{red}{TODO (#1):} \emph{#2}]}
\newcommand{\coq}{Coq\xspace}
\newcommand{\coqdoc}{Coqdoc\xspace}
\newcommand{\coqmakefile}{\texttt{coq\_makefile}\xspace}
\newcommand{\community}{Coq-community\xspace}
\newcommand{\gaia}{Gaia\xspace}
\newcommand{\alectr}{Alectryon\xspace}
\newcommand{\equations}{Equations\xspace}
\newcommand{\Hydras}{Hydras \& Co$\text.$\xspace}
\newcommand{\make}{\texttt{make}\xspace}

\definecolor{orange}{rgb}{1.0,0.6,0.0}
\definecolor{darkorange}{rgb}{0.7,0.25,0.0}
\definecolor{plugincolor}{rgb}{0.6,0.0,0.6}
\definecolor{mathcolor}{rgb}{0.0,0.0,0.6}
\definecolor{coqstylecolor}{rgb}{0.0,0.0,01.0}
\definecolor{lightblue}{rgb}{0.2,0.2,1.0}
\definecolor{metavarcolor}{rgb}{0.5,0.0,1.0}
\definecolor{darkgreen}{rgb}{0.1,0.7,0.1}
\definecolor{answercolor}{rgb}{.08,.15,.8}
\definecolor{normalcolor}{rgb}{0.0,0.0,0.0}
\definecolor{exbluecolor}{rgb}{0.1,0.1,0.9}
\definecolor{dontlookcolor}{rgb}{0.5,0.5,0.5}
\definecolor{termcolor}{rgb}{0.0,0.1,0.9}
\definecolor{lookcolor}{rgb}{0.5,0.1,0.0}
\definecolor{prooftermcolor}{rgb}{0.3,0.1,1.0}
\definecolor{typecolor}{rgb}{1.0,0.6,0.0}
\definecolor{taccolor}{rgb}{0.70,0.10,0.0}
\definecolor{pink}{rgb}{0.8,0.6,0.6}
\definecolor{darkmagenta}{rgb}{0.4,0.0,0.6}
\definecolor{darkblue}{rgb}{0.0,0.0,0.6}
\definecolor{cyan}{rgb}{0.0,0.4,0.8}
\usepackage{tikz}
\usepackage{tikzsymbols}
\usepackage{pifont}
\usetikzlibrary{arrows}

\usepackage{ifpdf}
\ifpdf
\usepackage{graphicx}
\else
\usepackage[dvips]{graphicx}
\fi

%%%% For Alectryon

\usepackage{texments}
\usepackage{./alectryon}
\usepackage{../pygments}

% Prevent breaks in the middle of syntactic units
\let\OldPY\PY
\def\PY#1#2{\OldPY{#1}{\mbox{#2}}}

%%% One hypothesis per line 
\makeatletter
\renewcommand{\alectryon@hyps@sep}{\alectryon@nl}
\makeatother

%%% \snippets{A,B,C,…} inputs a series of snippets as one block (with \itemsep
%%% between them).  A, B, C should be paths to files in snippets/.

\usepackage{etoolbox}
\makeatletter

\newcommand{\pathtomovies}{.}

\newcommand{\inputsnippets}[1]
{{\setlength{\itemsep}{1pt}\setlength{\parsep}{0pt}% Adjust spacing
    \alectryon@copymacros\begin{io}
      \forcsvlist{\item\@inputsnippet}{#1}
    \end{io}}}
\let\input@old\input % Save definition of \input
\newcommand{\@inputsnippet}[1]
{{\renewenvironment{alectryon}{}{}% Skip \begin{alectryon} included in snippet
    \input@old{{\pathtomovies}/#1}}}
\makeatother

% End of Alectryon stuff
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%% Specific macros

\newcommand{\canonseq}[2]{\mbox{$\{#1\}(#2)$}}
\newcommand{\rounds}{\mbox{\,\texttt{-+->}\,}}
\newcommand{\round}{\mbox{\,\texttt{-1->}\,}}
\newcommand{\rplus}[1]{\mbox{$\,\underset{#1}{\xrightarrow{\textcolor{white}{\,#1\,}}}\,$}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
  \frametitle{The \Hydras project}
  \begin{block}{}
    \begin{itemize}
   \item Experimental platform for the  development of \textcolor{lookcolor}{maintained} and \textcolor{lookcolor}{documented} libraries of formal proofs.
    \item A project of \community, developped on Github
      \begin{itemize}
      \item CI/CD
        \item Tools for collaboration (issues, PR, etc.)
        \end{itemize}
           \item book in pdf (300p), generated  with \textcolor{lookcolor}{\alectr} and \LaTeX, continuously updated and deployed
    \end{itemize}
  \end{block}
  \begin{block}{Contents}
    \begin{itemize}
    \item Evolved version  of two contributions: \textcolor{plugincolor}{Cantor} and \textcolor{plugincolor}{Additions}
    \item Maintenance of \textcolor{plugincolor}{Ackermann}
      (part of Russel O'Connor's \textcolor{plugincolor}{Goedel})
    \item Bridge to \textcolor{plugincolor}{Gaia} (José Grimm) (\emph{in construction})

@@

\begin{frame}
   \begin{block}{Example-driven books}
    \begin{itemize}
    \item The structure is determined by the presented topics (discrete math, algorithmics, etc.). \coq technology (formalization and proof patterns) is presented when needed.
    \item No need to be complete (there is already a reference manual)
    \item No need to introduce \coq (there are already a lot of initiation books and tutorials)
    \item Allow to compare several different formalizations of a given notion, and several proofs of the same theorem.
    \end{itemize}
  \end{block}
  \begin{block}{Examples}
    \begin{itemize}
    \item Tsukuba Coq Users's Group: regular languages, $\lambda$-calculus, arithmetic, ... (in Japanese + ssreflect)
    \item Software Foundations (B. Pierce \emph{et al.})
    \end{itemize}
  \end{block}
\end{frame}

@@

    % \item We want to ``compute'' the length of sequences of the form 
    %   \textcolor{mathcolor}{$\langle \alpha_k=\alpha,\dots,\alpha_{i+1}=\canonseq{\alpha_i}{i+1},\dots, 1,0\rangle$}, where \textcolor{mathcolor}{$k\in\mathbb{N}$}.

\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}

@@

@@

\begin{frame}
  \begin{block}{}
    \begin{itemize}
    \item Let \textcolor{mathcolor}{$0<\alpha<\epsilon_0$}, and \textcolor{mathcolor}{$i\in\mathbb{N}$}.
      The least index \textcolor{mathcolor}{$k$} such that \textcolor{mathcolor}{$\alpha_k=0$} is  \textcolor{mathcolor}{$H'_\alpha(i)$} where \textcolor{mathcolor}{$H'_\alpha$} is defined by transfinite induction over \textcolor{mathcolor}{$\alpha$}.

@@

  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \begin{frame}
    \frametitle{\community}
    \begin{block}{}
      \community is an informal organization that aims to maintain interesting open source \coq projects and facilitate collaboration among Coq users on documentation, tooling, etc.
      \begin{itemize}
      \item Created in 2018,  run by volunteer Coq users on GitHub
      \item At present, over than 50 projects, maintained by over than 30 volunteers
      \item Some packages were unmaintained before their integration to \community. According to their scientific or historical interest, we decided to revive them.
        \item After their transfer, packages may be the object of large changes  and refactorings.
      \end{itemize}
    \end{block}
  \end{frame}

>>>>

\documentclass[10pt]{beamer}
\usetheme{Montpellier}

\input{macros}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\begin{frame}
  \tableofcontents
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
    \frametitle{\community}
    \begin{block}{}
      \community is an informal organization that aims to maintain interesting open source \coq projects and facilitate collaboration among Coq users on documentation, tooling, etc.
      \begin{itemize}
      \item Created in 2018,  run by volunteer Coq users on GitHub
      \item At present, over than 50 projects, maintained by over than 30 volunteers
      \item Some packages were unmaintained before their integration to \community. According to their scientific or historical interest, we decided to revive them.
        \item After their transfer, packages may be the object of large changes  and refactorings.
      \end{itemize}
    \end{block}
  \end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
  \frametitle{The \Hydras project}
  \begin{block}{}
    \begin{itemize}
   \item Experimental platform for the  development of \textcolor{lookcolor}{maintained} and \textcolor{lookcolor}{documented} libraries of formal proofs.
    \item A project of \community, developped on Github
      \begin{itemize}
      \item CI/CD
        \item Tools for collaboration (issues, PR, etc.)
        \end{itemize}
           \item book in pdf (300p), generated  with \textcolor{lookcolor}{\alectr} and \LaTeX, continuously updated and deployed
    \end{itemize}
  \end{block}
  \begin{block}{Contents}
    \begin{itemize}
    \item Evolved version  of two contributions: \textcolor{plugincolor}{Cantor} and \textcolor{plugincolor}{Additions}
    \item Maintenance of \textcolor{plugincolor}{Ackermann}
      (part of Russel O'Connor's \textcolor{plugincolor}{Goedel})
    \item Bridge to \textcolor{plugincolor}{Gaia} (José Grimm) (\emph{in construction})

@@

\begin{frame}
   \begin{block}{Example-driven books}
    \begin{itemize}
    \item The structure is determined by the presented topics (discrete math, algorithmics, etc.). \coq technology (formalization and proof patterns) is presented when needed.
    \item No need to be complete (there is already a reference manual)
    \item No need to introduce \coq (there are already a lot of initiation books and tutorials)
    \item Allow to compare several different formalizations of a given notion, and several proofs of the same theorem.
      \item Show also interesting dead ends (formalization, tactics, proof structure).
    \end{itemize}
  \end{block}
  % \begin{block}{Examples}
  %   \begin{itemize}
  %   \item Tsukuba Coq Users's Group: regular languages, $\lambda$-calculus, arithmetic, ... (in Japanese + ssreflect)
  %   \item Software Foundations (B. Pierce \emph{et al.})
  %   \end{itemize}
  % \end{block}
\end{frame}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \begin{frame}
    \frametitle{Paths and accessibility}
    \begin{block}{Theorem (K. \& S., 1981) }
      % If $\beta<\alpha<\epsilon_0$, there exists $i$ and $l$ in $\mathbb{N}$ such that $ \alpha \rplus{i, i+1, i+2, \dots,l} \beta$.

    \begin{tikzpicture}[very thick, scale=0.25]
\node (alpha) at (0,0) {$\alpha$};
    \node (beta) at (40, 0){$\beta$};

  \draw[->, very thick,black] (alpha)--  node [above]{$>$} (beta);

  \node (alpha1) at (5,5) {};
  \node (alpha2) at (13,5) {};
    \node (alpha3) at (20,5) {};
  \node (alpha4) at (25,5) {};
  \draw [->, dashed,very thick,red] (alpha)-- node [below, rotate=40]{$i$}  (alpha1);
  \draw [->, dashed,very thick,red] (alpha1)-- node [below]{$i+1$}  (alpha2);
   \draw [->, dashed,very thick,red] (alpha2)-- node [below]{$i+2$}  (alpha3);
    \draw [->, dashed,very thick,red] (alpha3)-- node [below]{$i+3$}  (alpha4);
  \node (dots) at (30,5) {$\dots$};
  \node (search) at (32,5){};
 \draw [->, dashed, very thick,red] (search)-- node [below, rotate=-26]{$l$}  (beta);
\end{tikzpicture}

\pause
\textbf{Remark:} The proof is not trivial. The built path could miss $\beta$!

\vspace{4pt}

\textbf{Cor:} There is no decreasing measure mapping hydras to  $[0, \mu)$ (for any $\mu<\epsilon_0$)  for proving the termination of the considered kind of hydra battles.

@@

\begin{frame}
  \begin{block}{}
    \begin{itemize}
    \item Let \textcolor{mathcolor}{$0<\alpha<\epsilon_0$}, and \textcolor{mathcolor}{$i\in\mathbb{N}$}.
      The least index \textcolor{mathcolor}{$l$} such that \textcolor{mathcolor}{$\alpha_l=0$} is  \textcolor{mathcolor}{$H'_\alpha(i)$} where \textcolor{mathcolor}{$H'_\alpha$} is defined by transfinite induction over \textcolor{mathcolor}{$\alpha$}.

@@
